---
title: "Day 3: `uwIntroStats` introduction"
subtitle: "EPI/BIOST Bootcamp 2016"
author: "Brian D. Williamson"
date: "27 September, 2016"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## `uwIntroStats`

- `uwIntroStats` is an R package developed by: Scott Emerson, M.D. Ph.D.; Andrew J. Spieker, Ph.D.; myself (Brian D. Williamson); and Travis Hee Wai

- The package aims to make adopting R easier by streamlining presentation of output and making functions easier to work with

## Why `uwIntroStats`?

- Prints output in a more intuitive manner than base R
- Prints output that *we* think is important
- Wide array of functions for descriptive statistics
- Uses robust standard error estimates in all calculations

## Some useful functions
|Desired output       | STATA                 | R (base)              | R (`uwIntroStats`)
| --- | --- | --- | --- |
| read in data (.txt file) | `infile` | `read.table()` |
| read in data (.csv file) | `infile` | `read.csv()` | 
| mean | `summarize` | `summary()` | `descrip()` 
| | `mean()` | |
| std dev | `summarize` | `summary()` | `descrip()` 
|         |             | `sd()` | 
| variance | `summarize` | `var()` | `descrip()` 
| chi-square | `tabulate` | `chisq.test()` | `tabulate()` 

                         
## Some useful functions (continued)
|Desired output       | STATA                 | R (base)              | R (`uwIntroStats`)|
| --- | --- | --- | --- |
|boxplot              | `graph box`           | `boxplot()`           | `bplot()`         |
|histogram            | `histogram`           | `hist()`              |                   |
|density curve        | `kdensity`            | `plot(density())`     |                   |
|scatter plot         | `scatter`             | `plot()`              | `scatter()`       |
|correlation          | `correlate`           | `cor()` (correlation) | `correlate`       |
|                     |                       | `cov()` (covariance)  |                   |
|probability          | `display normprob(1)` | `1 - pnorm(1)`        |                   |
|(normal, z-score 1)  |                       |                       |                   |
|t-test               | `ttest`               | `t.test()`            | `ttest()`         |
|linear regression    | `regress`             | `lm()`                | `regress()`       |

## Example: `mri` data
- First, set your working directory. In RStudio, go to the pane with "Files", "Plots", etc. Then click "More", then "Set as working directory". Your console should print something like `setwd("~/Workshop2016-master/day_3_session_1")`
- Load the `mri` data:
```{r, echo = TRUE}
## load the package
library(uwIntroStats)

## load the data
data(mri)
```

## Example: `mri` data
- We can get descriptive statistics for the whole dataset using `descrip()`...
```{r, echo = TRUE}
descrip(mri)
```

## Example: `mri` data
- ... but this is unwieldy
- Easier to look at each variable individually, or a couple at a time (we will cover how to access single variables later)
```{r, echo = TRUE}
descrip(mri$age)
```
- We can also look within strata
```{r, echo = TRUE}
descrip(mri$age, strata = mri$male)
```

## Example: `mri` data
- Scatterplots help get a look at trends in the data

```{r, echo = TRUE}
scatter(mri$atrophy, mri$age, main = "Cerebral Atrophy vs Age", ylab = "Cerebral Atrophy", xlab = "Age (years)")
```

## Example: `mri` data
- Can stratify, e.g. based on sex

```{r, echo = TRUE}
scatter(mri$atrophy, mri$age, strata = mri$male, main = "Cerebral Atrophy vs Age, by Sex", ylab = "Cerebral Atrophy", xlab = "Age (years)")

```

## Summary
- `uwIntroStats` is an R package designed to make using R easier
- Provides flexible functions for descriptive statistics
- Many parallels with base R and with STATA