---
title: 'Day 3: Data types and data structures'
subtitle: "EPI/BIOST Bootcamp 2016"
author: "Brian D. Williamson"
date: "27 September, 2016"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Outline
- Data types
- Data structures
- Assigning values to variables
- Returning values from data structures

## Data types
- How is data stored? What type of data are we working with?
- A *data type* is a basic building block, and tells us how the data is stored
- R has three basic data types: numeric, character, and logical
- R also has two more complex data types: special numerics and logicals, and factors

## Numeric
- Decimal numbers
- Two types: 
    - Integers - whole numbers, e.g. 1
    - Doubles - real numbers (a whole number plus a fractional part), e.g. 1.5
- R designates these two as `int` and `num`
- Example: `age` variable in the `mri` data from the `uwIntroStats` package
```{r, message = FALSE, warning = FALSE}
## load the uwIntroStats package
library(uwIntroStats)

## load the mri data
data(mri)

## check the storage type of age
str(mri$age)
```
- Example: `weight` variable in the `mri` data from the `uwIntroStats` package
```{r}
str(mri$weight)
```

## Character
- Strings: words and letters, e.g. "Hello World!"
- R designates these as `chr`
- Example: "Mary had a little lamb" 
```{r}
str("Mary had a little lamb")
```
- Example: "A", "B", "C", "D", "E"
```{r}
str(c("A", "B", "C", "D", "E"))
```

## Logical
- `TRUE` or `FALSE`
- R designates these as `logi`
- `TRUE` corresponds to 1, `FALSE` to 0
- Produced by comparisons - the standard operators are:

Operator | Meaning | Return
:---:|:---:|---:
\| | or | `TRUE` if any argument is `TRUE`
& | and | `TRUE` only if all arguments are `TRUE`
! | negation | `FALSE` if argument returns `TRUE`, `TRUE` else
== | equals | `TRUE` if the two arguments are equal

- Examples:
    - Test if 1 equals 2: 1 == 2
    - Test if 1 equals 2 or 2 equals 2: (1 == 2) | (2 == 2)
    - Opposite of 1 equals 2: !(1 == 2)
    - Test equality of strings: "A" == "B"
    
## Special numeric
- R's way to handle invalid mathematical operations
- Two special numerics:
    - `Inf`: infinity. Defined as any nonzero number divided by zero 
        - one way this can occur is division by a very small number that R rounds to zero!
    - `NaN`: not a number. Defined as 0/0

## Special logical
- R's way to handle missing data
- `NA`: missing data
- R is essentially saying "I don't know what is supposed to go here"

## The null object
- `NULL`: the null object
- Improperly defined functions or expressions can sometimes return `NULL`

## Factors
- Store character strings (names) and numerical values
- Sometimes useful for keeping data organized
- Example: `male`, coded as 1 for female and 2 for male
    - a factor has these numeric values assigned to "Female" and "Male", or "F" and "M"
    
## Functions to check data types
- The `str` function returns the data type of a variable
- The "Environment" tab has information like data type and size of arrays/matrices
- Example: `age` from the `mri` data
```{r, echo = TRUE}
str(mri$age)
```
- The class of `is.` functions perform logical comparisons
    - `is.numeric()` tests whether the variable is numeric or not
    - `is.character()` tests whether the variable is a character or not
    - `is.logical()` tests whether the variable is logical or not
    - `is.finite()` and `is.infinite()` test whether each value is finite or infinite
    - `is.na()` tests whether each value is `NA` or not
- Example: `age` from the `mri` data
```{r, echo = TRUE}
is.numeric(mri$age)
sum(is.na(mri$age))
```

## Exercise: data types
1. What is the data type of:
    1. 1?
    2. 2.5?
    3. `TRUE`
    4. "Seattle"
2. What do the following comparisons return?
    1. "A" == "B"
    2. ("A" == "A") & (1 == 1)
3. What does `is.na(1)` return?

## Solution: data types
1. Solutions are:
    1. Numeric, integer, `int`
    2. Numeric, double, `num`
    3. Logical, `logi`
    4. Character, `chr`
2. Solutions:
    1. `FALSE`
    2. `TRUE`
3. `FALSE`

## Data structures
- Equally important to what type of data we have is how it is stored
- *Data structures* are the various ways that R stores data
- We have been working with data structures implicitly already - each variable is part of a data structure
- Five basic data structures in R:
    - Vectors
    - Matrices
    - Arrays
    - Data frames
    - Lists

## Vectors
- Store any number of individual values
- All values must be the same data type
    - If not, R will force all values to be the same type (with perhaps varying results)
- Create vectors with `c()`
- Examples:
    - vector of strings: `c("Mary", "had", "a", "little", "lamb")`
    - vector of integers: `c(1, 2, 3, 4, 5)`
    
```{r, echo = TRUE}
## same string as before
c("Mary", "had", "a", "little", "lamb")

## trying two data types
c("Hello", "number", 1)
c(1, "2", 3)
```

## Matrices
- Tables, or two-dimensional vectors
- All values must be the same data type
- Must be rectangular: number of rows in each column is equal, and number of columns in each row is equal 
- Create matrices with `matrix(vector, nrow, ncol, byrow, dimnames)`
    1. Create a vector inside this call with `c()`
    2. Specify how many rows (`nrow`) and/or columns (`ncol`)
    3. Specify whether to read in values across rows (`byrow = TRUE`) or down columns (`FALSE`)
    4. (optional) Specify the names of the dimensions, in a list

## Matrices: examples
```{r, echo = TRUE}
## create a matrix from a vector of numbers
matrix(c(1, 0, 0, 1), nrow = 2, ncol = 2, byrow = TRUE)

## matrix from a vector of words (matrices can be non-square)
matrix(c("This", "is", "a", "matrix"), nrow = 1, byrow = TRUE)
```

## Matrices: when might we use them?
- Storing data of one type (like all numerics)
- Tables of output
- Some functions only take matrices as input (see documentation)
- My default data structure

## Arrays
- Higher-order matrices (a matrix is also a 2-D array)
- All values must be the same data type
- Must be rectangular, like matrices
- Create with `array(vector, dim)`
    1. Create a vector with `c()`
    2. Use `dim` to specify an integer vector with the maximum number of indices in each dimension
    3. (optional) Specify the names of the dimensions, in a list

## Arrays: examples
```{r, echo = TRUE}
## 1-D array, or vector
array(c(1, 2, 3, 4), dim = 4)

## 2-D array, or matrix
array(c(1, 0, 0, 1), dim = c(2, 2))

## 3-D array (an array with two 2-D sub-arrays)
array(c(1, 0, 0, 1, 1, 0, 0, 1), dim = c(2, 2, 2))
```

## Arrays: when might we use them?
- High dimensional tables (stratification on multiple variables)

## Data frames
- A table, like a matrix
- Columns may have different data types
    - For this reason, data frames are a common storage mode in R
- Each column must have the same number of rows
- Create with `data.frame(..., row.names, stringsAsFactors)`
    1. `...` is a special object, meaning that we can put anything here: a comma separated list of vectors, matrices etc., or many other objects
    2. (optional) Specify the names of the columns using `=`, with the name on the left and the object on the right
    3. (optional) Specify the names of the rows using `row.names`
    4. (optional) Convert strings to factors if possible (`stringsAsFactors = TRUE`)
- The `names()` function returns all variable names

## Data frames: examples
```{r, echo = TRUE}
data.frame(c(1,2,3), c("UW", "is", "awesome")) ## data frame with no column names
data.frame(nums = c(1,2,3), words = c("UW", "is", "awesome")) ## data frame with column names
data.frame(nums = c(1,2,3), words = c("UW", "is", "awesome"), mat = matrix(c(1, 0, 0, 0, 0, 1), nrow = 3)) ## input vectors and matrices
```

## Data frames: when might we use them?
- Storing multiple data types in one object
- Some functions only take data frames as input (like `lm()`; see documentation for others)
- Want to name the columns and reference them using `$`

## Lists
- An ordered collection of objects
- Each object can have a different data type, or can be any data structure
- Created by `list(...)`
    - Again the `...`! So lots of things can go here!
    - We can also name objects using `=`, like in `data.frame()`
    
## Lists: examples
```{r, echo = TRUE}
list(1, "UW", 2, "is", 3, "awesome") ## list with single characters and strings
list(df = data.frame(nums = c(1,2,3), words = c("UW", "is", "awesome")), mat = matrix(c(1, 0, 0, 0, 0, 1), nrow = 3)) ## list with named data frame and matrix
```

## Lists: when might we use them?
- Store many different data types or structures in one object
- Name each object and reference it easily
- Honestly my least often used data structure

## Exercise: data structures
1. How does one create a vector?
2. What will happen when we enter `c(1, 2, 3, 4, "a")`?
3. Create the matrix

```{r}
matrix(c(1, 2, 3, 4, 5, 6), nrow = 2, byrow = TRUE)
```

4. Can each column have a different data type in a data frame?

## Solution: data structures
1. `c()` creates a vector
2. The numeric values will be coerced into character values, yielding 

```{r, echo = TRUE}
c(1, 2, 3, 4, "a")
```

3. 
```{r, echo = TRUE}
matrix(c(1, 2, 3, 4, 5, 6), nrow = 2, byrow = TRUE) # this is one way, specifying ncol = 3 works too
```

4. Yes!!

## Assigning values to variables
- So far, we have not been saving values - we have just been looking at them in the console
- Saving values makes manipulation easier
- The assignment operator, `<-`, allows us to save values to objects/variables
- Choosing names that we describe what the object is will save us time later on!
- Examples:
```{r, echo = TRUE}
junk <- 1 ## create an object called "junk", assign it value of 1
vec <- c(1, 2, 3, 4, 5) ## create a vector called "vec", assign it the vector 1-5
mat <- matrix(c(1, 0, 0, 1), ncol = 2, byrow = TRUE) ## matrix called "mat"

junk ## prints out the value of junk
vec ## prints out the value of vec
mat ## prints out the value of mat
```

## Returning a single value from a data structure
- Three main ways to access data
- These deal with data in vector or matrix form, list form, and data frame form
- Accessing data in vectors, matrices, and data frames is most simple

## The `[]` operator
- Allows us to access values in vectors, matrices, or data frames; use depends on the data structure
- For vectors, we access one or multiple values using a single index or a vector of indices, e.g. `c(1, 2, 3)`
- For matrices or data frames, we can access values in rows or columns, using `[row indices, col indices]`
- Examples:
```{r, echo = TRUE}
vec[1] ## get the first element from vec
vec[c(1, 3, 5)] ## get the first, third, and fifth elements from vec
vec[2:4] ## shortcut to get the second, third, and fourth elements from vec
mat[1, 1] ## the element in the first row and first column of mat
mat[1, ] ## all columns of the first row of mat
```

## The `$` operator
- Allows us to extract columns from a data frame
- Once we extract a column, we can access values using `[]`
- Examples:
```{r, echo = TRUE}
df <- data.frame(nums = c(1,2,3), words = c("UW", "is", "awesome")) ## create a data frame with two columns
df$nums[1] ## access the first value in the "nums" column
df[1, "nums"] ## another way to get the first value in the "nums" column
df[1, 1] ## another way to get the first value in the "nums" column
df$words[3] ## access the third value in the "words" column
```

## Indexing values in a data structure
- Accessing more than one value is often useful
- As easy as accessing a single value!

## The `[]` operator: revisited
- Returning multiple elements from a vector: give `[]` a vector of indices to return, e.g. `vec[2:4]`
- Returning multiple elements from a matrix:
    - For entire rows, give a vector of row indices with no column indices, e.g. `mat[1, ]`
    - For entire columns, give a vector of column indices with no row indices, e.g. `mat[, 1]`
    - Only selecting one row or column from a matrix returns a vector by default (there are ways around this)
- Examples:
```{r, echo = TRUE}
mat.2 <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 2, byrow = TRUE)
mat.2[1, ] ## return the first row from mat.2
mat.2[, c(2,3)] ## return the second and third columns from mat
```

## The `$` operator: revisited
- `$` works with data frames and lists
- Data frames: 
    - We can only access one variable at a time using `$`
    - However, we can access multiple values within each variable (or column) using vectors of indices
- Lists:
    - `$` returns one object from the list, as long as that object is named
    - We can then access multiple values using the rules for that object
- Examples:
```{r, echo = TRUE}
df$nums ## access all values from the "nums" column of df
lst <- list(df = data.frame(nums = c(1,2,3), words = c("UW", "is", "awesome")), mat = matrix(c(1, 0, 0, 0, 0, 1), nrow = 3)) ## list with named data frame and matrix
lst$df ## return the data frame from lst
lst$df[1, 1] ## first element from df
lst$df$nums ## "nums" variable from df
```

## The `[[]]` operator
- Access objects in a list
- Use `[[#]]` to extract the \# th object
- Examples:
```{r, echo = TRUE}
lst[[1]] ## first object in lst
lst[[2]][1, ] ## first row from the second object in lst
```
## Advanced indexing
- So far we have only indexed using vectors or single indices
- Other types of indexing:
    - Negative numbers or vectors of indices: exclude elements
    - Logicals: include only elements for which the expression is `TRUE`
- Examples:
```{r, echo = TRUE}
vec[-1] ## all elements from vec except the first
mat[-1, ] ## all elements except the first row of mat
df[-c(1, 2), ] ## all elements except the first and second rows of df
df[df$nums == 1, ] ## all rows from df where df$nums equals 1
```

## Exercise: accessing values and indexing
For these exercises, load the `mri` data from `uwIntroStats` using `data(mri)`

1. Return the `age` variable from this dataset
2. Return the first five values from the `age` variable
3. Count the number of people greater than 75 years old (Hint: use logicals)
4. For those people less than 75 years old, return their height, weight, and sex

## Solution: accessing values and indexing
```{r, echo = TRUE}
## load the data
data(mri)

## 1. Return the age variable. Note: I am using head() to display only the first 6 values to save space
head(mri$age) ## alternatively, could use mri[, "age"] or mri[, 3]

## 2. first five values from age
mri$age[1:5]

## 3. Number of people > 75 years old
sum(mri$age > 75) ## alternatively, could use dim(mri[mri$age > 75, ])

## 4. Return height, weight, sex for those > 75 (again using head() to save space)
head(mri[mri$age > 75, c("height", "weight", "male")])
```

## Summary
- R has five data types: numeric, character, logical, factor, special numerics and logicals
- Each data type stores different information
- R has five basic data structures: vector, matrix, array, data frame, list
- Some of these can only contain one data type, others can contain multiple
- We can access data using vectors of indices, logicals, or both
- The operator we use depends on the data structure (`[]`, `$`, or `[[]]`)